Quasi-Monte Carlo Methods: What, Why, and How?

### 2025-01-08 HickernellMCQMC2024_QMCTutorial.tex
M2 Mac updates

\begin{abstract}
Many problems in  quantitative finance, uncertainty quantification, and other areas can be formulated as computing $\mu := \mathbb{E}(Y)$, where instances of $Y:=f(\boldsymbol{X})$ are generated by numerical simulation. The population mean, $\mu$, can be approximated by the sample mean, $\hat{\mu}_n := n^{-1} \sum_{i=1}^n f(\boldsymbol{X}_i)$.  Computing $\mu$ is equivalent to computing a $d$-dimensional integral.

Quasi-Monte Carlo methods replace independent and identically distributed  sequences of random vectors, $\{\boldsymbol{X}_1, \boldsymbol{X}_2, \ldots \}$, by low discrepancy sequences.  This accelerates the convergence of $\hat{\mu}_n$ to $\mu$ as $n \to \infty$.

This tutorial describes  low discrepancy sequences  and their quality measures.  We demonstrate the performance gains possible with quasi-Monte Carlo methods.  Moreover, we describe how to formulate problems to realize the most increase in performance using quasi-Monte Carlo methods.  We also briefly describe the use of quasi-Monte Carlo methods for problems beyond computing the mean.

\keywords{}
\end{abstract}
%
\section{Introduction} \label{sec:intro}
There are may settings where key underlying quantities that affect the outcome are unknown, e.g.,
\begin{itemize}
	\item Future market forces, which affect financial risk,
	\item The porosity of rock, which affect the extraction of oil or gas, or
	\item Elemetary particle interactions in a high energy physics experiment, which affect what is observed.
\end{itemize}
In such situations the unknown inputs are often modeled using random variables or stochastic processes.  Computations are performed by generating a multidude of possible scenarios informed by the assumed probability distribution of the input. These are used to estimate the mean, quantiles, and/or probability distribution of the outcome.  This is the essence of the Monte Carlo (MC) method.

In mathematical terms, the outcome is $Y := f(\boldsymbol{X})$, where $\boldsymbol{X}$ is a vector random variable of inputs determining the outcome.  The values of the function $f$ can be computed by an algorithm, whose complexity may be such that $f$ can be considered a ``black box''.  The scenarios are $\bsX_0, \bsX_1, \ldots$, which give rise to outcomes $Y_0 = f(\bsX_0), Y_1 = f(\bsX_1), \ldots$.  The $Y_i$ form the basis for approximating $\bbE(Y)$ and other quantities of interest.

Simple or independent and identically distributed (IID)  Monte Carlo chooses the sequence $\bsX_0, \bsX_1, \ldots$ to be IID.  Quasi-Monte Carlo (qMC) methods choose $\bsX_0, \bsX_1, \ldots$ \emph{not} IID, but with an empirical distribution that approximates the true probability distribution of $\bsX$.  The difference between these two distributions is called a \emph{discrepancy}, and the sequences used in qMC are called low discrepancy sequences.

This tutorial describes what qMC is, why we would want to use qMC, and how qMC can be implemented well.  The next session provides an illustration of qMC in practice.  This is followed by a description of how low discrepancy sequences are constructed and various measures of quality.  We then explain how randomization can improve qMC, how to decide what sample size, $n$, is sufficient, and how to rewrite the problem of interest in a qMC-friendly way.

@@

\begin{equation}\label{eq:keisterA}
	\mu := \int_{\mathbb{R}^d} \cos(\lVert \bst \rVert) \exp(-\lVert \bst \rVert^2) \, \rmd \bst,
\end{equation}
where $\lVert \bst \rVert := \sqrt{t_1^2 + \cdots + t_d^2}$.  This integral may be evaluated numerically by re-writing it in spherical coordinates as
\begin{equation}\label{eq:keisterExact}
	\mu = \frac{2 \pi^{d/2}}{\Gamma(d/2)}\int_{0}^{\infty} \cos(r) \exp(-r^2) \, r^{d-1} \rmd r,
\end{equation}
where $2 \pi^{d/2}/\Gamma(d/2)$ is the surface area of the sphere in $d$ dimensions, and $\Gamma$ is the Gamma function.  Having a non-trivial test case with a value that can be easily calculated allows us to compute the numerical errors of various cubature schemes.

The aim is approximate $\mu$ by a sample mean,
\begin{equation} \label{eq:sample_mean}
	\hat{\mu}_n := \frac 1n \sum_{i=1}^n f(\boldsymbol{X}_i)
\end{equation}
but to do so requires some further preparation.  This integral above, $\mu$, may be thought of as the expectation of $Y := g(\bsT) := \pi^{d/2} \cos(\lVert \bsT \rVert)$, where $T_1, \ldots, T_d$ are IID Gaussian random variables with zero mean and variance $1/2$, i.e., $\bsT \sim \calN(\bszero,\mathsf{I}/2)$:
\begin{equation}\label{eq:keisterB}
	\mu = \int_{\mathbb{R}^d} \underbrace{\pi^{d/2} \cos(\lVert \bst \rVert)}_{=:g(\bst)} \cdot \underbrace{\frac{\exp(-\lVert \bst \rVert^2)}{\pi^{d/2}}}_{\text{density of } \calN(\bszero,\mathsf{I}/2)} \, \rmd \bst = \bbE(Y) = \bbE[g(\bsT)].
\end{equation}

Since the low discrepancy sequences that drive qMC are defined on the unit cube, $[0,1]^d$, we perform a variable transformation $\bst = \bigl (\Phi^{-1}(x_1), \ldots, \Phi^{-1}(x_d) \bigr )/\sqrt{2}$, where is $\Phi$ is the cumulative distribution function of the standard Gaussian random variable.  This reimagines the integral $I$ as the expectation of a function, $f$, of a standard uniform random variable:
\begin{multline}\label{eq:keisterC}
	\mu = \int_{[0,1]^d} \underbrace{\pi^{d/2} \cos\Bigl (\sqrt{2} \bigl \lVert \bigl( \Phi^{-1}(x_1), \ldots, \Phi^{-1}(x_d)\bigr) \bigr\rVert \Bigr)}_{=:f(\bsx)} \, \rmd \bsx \\
	= \bbE[Y]
	= \bbE[f(\bsX)], \qquad \bsX \sim \calU[0,1]^d.
\end{multline}

Consider the specific case of $d=6$ for which $\mu \approx -2.327303729298$.  We approximate $\mu$ by the sample mean defined in \eqref{eq:sample_mean} for three different kinds of nodes, $\bsX_0, \bsX_1, \ldots, \bsX_{n-1}$, for various sample sizes, $n$, and plot the relative errors, $\lvert (\mu - 	\hat{\mu}_n)/\mu\rvert$ in Figure \ref{fig:keister-err}. The three kind of nodes are
\begin{enumerate}
	\renewcommand{\labelenumi}{\roman{enumi}.}
	\item Cartesian grids, $\{1/(2m), \ldots, (2m-1)/(2m) \}^6$, for $m = 2, 3, \ldots$ (blue),
	\item IID simple random  with arbitrary $n$ (orange), and
	\item Randomized low discrepancy (Sobol') sequences with $n = 1, 2, \ldots, 2^m, \ldots $ (green).
\end{enumerate}

@@

	\item \emph{Grid nodes.} Although grids may be attractive for low dimensional problems, e.g., $d = 1$, $2$, or $3$, they do poorly for this modest dimension, $d=6$.  For this modest $d$, the possible sample sizes, $n$, are quite sparse, as shown in Figure \ref{fig:keister-err}.

>>>>

\begin{abstract}
Many problems in  quantitative finance, uncertainty quantification, and other areas have answers of the form $\mu := \mathbb{E}(Y)$, where instances of $Y:=f(\boldsymbol{X})$ may be generated by numerical simulation. The population mean, $\mu$, can be approximated by the sample mean, $\hat{\mu}_n := n^{-1} \sum_{i=0}^{n-1} f(\boldsymbol{X}_i)$ for a well chosen sequence of nodes, $\{\bsX_0, \bsX_1, \ldots\}$.  Computing $\mu$ is equivalent to computing a $d$-dimensional integral.

Quasi-Monte Carlo methods replace independent and identically distributed  sequences of random vector nodes, $\{\boldsymbol{X}_i \}$, by low discrepancy sequences.  This accelerates the convergence of $\hat{\mu}_n$ to $\mu$ as $n \to \infty$.

This tutorial describes  low discrepancy sequences  and their quality measures.  We demonstrate the performance gains possible with quasi-Monte Carlo methods.  Moreover, we describe how to formulate problems to realize the most increase in performance using quasi-Monte Carlo methods.  We also briefly describe the use of quasi-Monte Carlo methods for problems beyond computing the mean, $\mu$.

\keywords{}
\end{abstract}
%
\section{Introduction} \label{sec:intro}
There are may settings where key underlying quantities that affect the outcome are unknown, e.g.,
\begin{itemize}
	\item Future market forces, which affect financial risk,
	\item The porosity of rock, which affect the extraction of oil or gas, or
	\item Elemetary particle interactions in a high energy physics experiment, which affect what is observed by detectors.
\end{itemize}
In such situations, the unknown inputs are often modeled using random variables or stochastic processes.  Computations are performed by generating a multidude of possible scenarios informed by the assumed probability distribution of the input. These are used to estimate the mean, quantiles, and/or probability distribution of the outcome.  This is the essence of the Monte Carlo (MC) method.

In mathematical terms, the outcome is $Y := f(\boldsymbol{X})$, where $\boldsymbol{X}$ is a vector random variable of inputs determining the outcome.  The values of $f$ can be computed by an algorithm, whose complexity may be such that $f$ can be considered a ``black box''.  The scenarios are $\bsX_0, \bsX_1, \ldots$, which give rise to outcomes $Y_0 = f(\bsX_0), Y_1 = f(\bsX_1), \ldots$.  The $Y_i$ form the basis for approximating $\bbE(Y)$ and other quantities of interest. These scenarios are also commonly called instances of $\bsX$ or nodes.  We use these terms interchangeably.

Simple or independent and identically distributed (IID)  Monte Carlo chooses the sequence of nodes $\{\bsX_0, \bsX_1, \ldots\}$ to be IID.  Quasi-Monte Carlo (qMC) methods choose the nodes to be \emph{not} IID, but with an empirical distribution that approximates well the true probability distribution of $\bsX$.  The difference between these two distributions is called a \emph{discrepancy}, and the node sequences used in qMC are called low discrepancy sequences.

This tutorial describes what qMC is, why we would want to use qMC, and how qMC can be implemented well.  The next session provides an illustrative example of the advantages of qMC.  This is followed by a description of how low discrepancy sequences are constructed and various measures of discrepancy.  We then explain how randomization can improve qMC, how to decide what sample size, $n$, is sufficient, and how to rewrite the problem of interest in a qMC-friendly way.

@@

\begin{equation}\label{eq:keisterA}
	\mu := \int_{\mathbb{R}^d} \cos(\lVert \bst \rVert) \exp(-\lVert \bst \rVert^2) \, \rmd \bst,
\end{equation}
where $\lVert \bst \rVert := \sqrt{t_1^2 + \cdots + t_d^2}$.  This integral may be evaluated numerically by re-writing it in spherical coordinates as
\begin{equation}\label{eq:keisterExact}
	\mu = \frac{2 \pi^{d/2}}{\Gamma(d/2)}\int_{0}^{\infty} \cos(r) \exp(-r^2) \, r^{d-1} \rmd r,
\end{equation}
where $2 \pi^{d/2}/\Gamma(d/2)$ is the surface area of the sphere in $d$ dimensions, and $\Gamma$ is the Gamma function.  This non-trivial test case with a true value that can be easily calculated allows us to compute the numerical errors of various cubature schemes.

The aim is to approximate $\mu$ by a sample mean,
\begin{equation} \label{eq:sample_mean}
	\hat{\mu}_n := \frac 1n \sum_{i=1}^n f(\boldsymbol{X}_i),
\end{equation}
but to do so requires some further preparation.  This integral above, $\mu$, may be thought of as the expectation of $Y := g(\bsT) := \pi^{d/2} \cos(\lVert \bsT \rVert)$, where $T_1, \ldots, T_d$ are IID Gaussian random variables with zero mean and variance $1/2$, i.e., $\bsT :=(T_1, \ldots, T_d) \sim \calN(\bszero,\mathsf{I}/2)$:
\begin{equation}\label{eq:keisterB}
	\mu = \int_{\mathbb{R}^d} \underbrace{\pi^{d/2} \cos(\lVert \bst \rVert)}_{=:g(\bst)} \cdot \underbrace{\frac{\exp(-\lVert \bst \rVert^2)}{\pi^{d/2}}}_{\text{density of } \calN(\bszero,\mathsf{I}/2)} \, \rmd \bst = \bbE(Y) = \bbE[g(\bsT)].
\end{equation}

Since the low discrepancy sequences that underly qMC are defined to approximate the standard uniform distribution, we perform a variable transformation $\bst = \bigl (\Phi^{-1}(x_1), \ldots, \Phi^{-1}(x_d) \bigr )/\sqrt{2}$, where is $\Phi$ is the cumulative distribution function of the standard Gaussian random variable.  This reimagines the integral $\mu$ as the expectation of a function, $f$, of a standard uniform random variable:
\begin{multline}\label{eq:keisterC}
	\mu = \int_{[0,1]^d} \underbrace{\pi^{d/2} \cos\Bigl (\bigl \lVert \bigl( \Phi^{-1}(x_1), \ldots, \Phi^{-1}(x_d)\bigr) \bigr\rVert/\sqrt{2}  \Bigr)}_{=:f(\bsx)} \, \rmd \bsx \\
	= \bbE[Y]
	= \bbE[f(\bsX)], \qquad \bsX \sim \calU[0,1]^d.
\end{multline}

Consider the specific case of $d=6$ for which $\mu \approx -2.327303729298$.  We approximate $\mu$ by the sample mean defined in \eqref{eq:sample_mean} for three different kinds of nodes, $\{\bsX_i\}_{i=0}^{n-1}$, for various sample sizes, $n$, and plot the relative errors, $\lvert (\mu - 	\hat{\mu}_n)/\mu\rvert$ in Figure \ref{fig:keister-err}. The three kind of nodes are
\begin{enumerate}
	\renewcommand{\labelenumi}{\roman{enumi}.}
	\item Cartesian grids, $\{1/(2m), \ldots, (2m-1)/(2m) \}^6$, for $m = 2, 3, \ldots$ (blue),
	\item IID simple random  with arbitrary $n$ (orange), and
	\item Randomized low discrepancy (Sobol') sequences with $n = 1, 2, \ldots, 2^m, \ldots $ (green).
\end{enumerate}

@@

	\item \emph{Grid nodes.} Although grids may be attractive for low dimensional problems, e.g., $d = 1$, $2$, or $3$, they do poorly for this modest dimension, $d=6$.  Figure \ref{} shows a grid with $n = 2$, $3$, and $6$For this modest $d$, the possible sample sizes, $n$, are quite sparse, as shown in Figure \ref{fig:keister-err}.
